---
title: "Practicum_scratch"
author: "Rebekah Adams"
date: "2023-01-16"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Set up
```{r packages}
#install.packages("tidyverse")
#install.packages("sf")
#install.packages("tidycensus")
#install.packages("riem")
#install.packages("gridExtra")
library(tidyverse)
library(sf)
library(tidycensus)
library(riem)
library(gridExtra)

plotTheme <- theme(
  plot.title =element_text(size=12),
  plot.subtitle = element_text(size=8),
  plot.caption = element_text(size = 6),
  axis.text.x = element_text(size = 10, angle = 45, hjust = 1),
  axis.text.y = element_text(size = 10),
  axis.title.y = element_text(size = 10),
  # Set the entire chart region to blank
  panel.background=element_blank(),
  plot.background=element_blank(),
  #panel.border=element_rect(colour="#F0F0F0"),
  # Format the grid
  panel.grid.major=element_line(colour="#D0D0D0",size=.2),
  axis.ticks=element_blank())

mapTheme <- theme(plot.title =element_text(size=12),
                  plot.subtitle = element_text(size=8),
                  plot.caption = element_text(size = 6),
                  axis.line=element_blank(),
                  axis.text.x=element_blank(),
                  axis.text.y=element_blank(),
                  axis.ticks=element_blank(),
                  axis.title.x=element_blank(),
                  axis.title.y=element_blank(),
                  panel.background=element_blank(),
                  panel.border=element_blank(),
                  panel.grid.major=element_line(colour = 'transparent'),
                  panel.grid.minor=element_blank(),
                  legend.direction = "vertical", 
                  legend.position = "right",
                  plot.margin = margin(1, 1, 1, 1, 'cm'),
                  legend.key.height = unit(1, "cm"), legend.key.width = unit(0.2, "cm"))
```

Read in data
```{r}
# read in quarterly files
q1_2022 <- st_read("./data/indego-trips-2022-q1.csv")
q2_2022 <- st_read("./data/indego-trips-2022-q2.csv")
q3_2022 <- st_read("./data/indego-trips-2022-q3.csv")
q4_2022 <- st_read("./data/indego-trips-2022-q4.csv")
```

Combine data into annual dataframes

**Question for Michael and Matt**: A single year has 900k+ rows - caching, processing, and rendering a single year's data is going to be tough to compute, let alone eight years' worth of data. What's a good alternative?
```{r}
# rbind
data_2022 <- rbind(q1_2022, q2_2022, q3_2022, q4_2022)

#900k rows - how best to run things without crashing my computer?
```

## Stations
Group data into stations, georeference
```{r stations}
stations = data_2022 %>%
  group_by(start_station) %>%
  summarize(stn_lat = median(as.double(start_lat), na.rm = T),
            stn_lon = median(as.double(start_lon), na.rm = T),
            created = min(start_time), #trying to idenfity when the station opened - this doesn't really matter until we have all years' data
            closed = max(start_time), #trying to identify if/when the station closed - this doesn't really matter until we have all years' data
            ridership = length(unique(trip_id))) %>% 
  na.omit(.) %>%
  st_as_sf(coords = c(x="stn_lon", y="stn_lat"), crs=st_crs("EPSG:4326"))

glimpse(stations)

```

## Data Sources
### Census data
Pull census data for PHL's blockgroups
```{r census data, results = FALSE}
census_api_key("INSERT_KEY_HERE", overwrite = TRUE)

#Call the census to view a list of available variables:
acs_variable_list.2020 <- load_variables(2020, #year
                                         "acs5", #five year ACS estimates
                                         cache = TRUE)

acs_vars <- c("B01001_001E", #ACS total Pop estimate
                   "B02001_002E", #Estimated total White only population
                   #"B08013_001E", #Estimated Aggregate travel time to work (in minutes)
                   #"B08012_001E", #Estimated Total working population who commute
                   "B23025_004E", #Estimated Total population who are employed
                   "B25058_001E", #Estimated Median Rent
                   "B19013_001E") #Estimated Median Household Income

blockgrps <- get_acs(geography = "block group",
                             year = 2020, 
                             variables = acs_vars, 
                             geometry = TRUE, 
                             state = "PA", 
                             county = c("Philadelphia"),
                             output = "wide") %>%
  dplyr::select (GEOID, NAME, geometry, all_of(acs_vars)) %>%
  rename (total_pop = B01001_001E, #ACS total Pop estimate
          total_white = B02001_002E, #Estimated total White only population
          #total_commute_time = B08013_001E, #Estimated Aggregate travel time to work (in minutes)
          #total_commuters =  B08012_001E, #Estimated Total working population who commute
          total_employed = B23025_004E, #Total employed
          medRent = B25058_001E, #Estimated Median Rent
          medIncome = B19013_001E, #Median HH income
          ) %>% # total kids of middle/highschool age that attend public school
  mutate(pct_white = (total_white / total_pop),
         year = 2020,
         pct_employed = (total_employed / total_pop),
         #commuting_time = (total_commute_time / total_commuters),
         #pct_commuters = (total_commuters / total_pop),
         pct_rent_spent = ((medRent * 12) / medIncome) * 100) %>%
  #dplyr::select(-total_commute_time, -total_white, -total_employed, -total_commuters) %>%
  st_transform("EPSG:4326")

# Extract centroids of block groups - these will be our origins
blockgrps_centroids <- st_centroid(blockgrps) %>%
  st_sf() %>%
  st_transform(st_crs(blockgrps))
```

Join stations to their block group's census data
```{r join stations and block groups}

station_census <- st_join(stations, blockgrps,
                          join=st_intersects,
                          left = TRUE) %>%
  rename("station" = "start_station") #rename so we don't get confused later

glimpse(station_census)
```

Join ride data to start and end stations' census data
```{r rides geoids}
rides <- data_2022 %>%
  left_join(., station_census, by = c("start_station"="station")) %>%
  rename("start_GEOID" = "GEOID",
         "start_total_pop" = "total_pop",
         "start_medRent" = "medRent",
         "start_medIncome" = "medIncome",
         "start_pct_white" = "pct_white",
         "start_pct_employed" = "pct_employed",
         "start_pct_rent_spent" = "pct_rent_spent") %>%
  dplyr::select(-created, -closed, -NAME, -total_white, -total_employed, -year, -ridership) %>%
  st_drop_geometry() %>%
  left_join(., station_census, by = c("end_station"="station")) %>%
  rename("end_GEOID" = "GEOID",
         "end_total_pop" = "total_pop",
         "end_medRent" = "medRent",
         "end_medIncome" = "medIncome",
         "end_pct_white" = "pct_white",
         "end_pct_employed" = "pct_employed",
         "end_pct_rent_spent" = "pct_rent_spent") %>%
  dplyr::select(-created, -closed, -NAME, -total_white, -total_employed, -year, -ridership) %>%
  mutate(start_time = as.POSIXct(start_time, tz="", format = "%m/%d/%Y %H:%M"),
         end_time = as.POSIXct(end_time, tz="", format = "%m/%d/%Y %H:%M"),
         interval60 = ymd_h(substr(start_time,1,13)),
         week = week(interval60),
         dotw = wday(interval60, label=TRUE),
         duration = as.numeric(duration))
         

glimpse(rides)
```

### Time
```{r time}

```

### Weather
Weather will probably not ultimately be very helpful to us, since this isn't a temporal model - but I'm including it here to help us ID seasonal variation in ridership.
```{r weather}
weather.Panel <- 
  riem_measures(station = "PHL", date_start = "2022-01-01", date_end = "2022-12-31") %>%
  dplyr::select(valid, tmpf, p01i, sknt)%>%
  replace(is.na(.), 0) %>%
    mutate(interval60 = ymd_h(substr(valid,1,13))) %>%
    mutate(week = week(interval60),
           dotw = wday(interval60, label=TRUE)) %>%
    group_by(interval60) %>%
    summarize(Temperature = max(tmpf),
              Precipitation = sum(p01i),
              Wind_Speed = max(sknt)) %>%
    mutate(Temperature = ifelse(Temperature == 0, 42, Temperature))

glimpse(weather.Panel)
```

```{r plot weather}
weather_panel <- grid.arrange(
  ggplot(weather.Panel, aes(interval60,Precipitation, color = Precipitation)) + geom_line() + scale_color_gradient(low = "#d0d1e6", high = "#0570b0")+
  labs(title="Precipitation", x="Hour", y="inches") + theme_bw(),
  
  ggplot(weather.Panel, aes(interval60,Wind_Speed, color = Wind_Speed)) + geom_line() + 
    scale_color_gradient(low = "#9ebcda", high = "#4d004b")+
    labs(title="Wind Speed", x="Hour", y="mph", color="Wind Speed") + theme_bw(),
  
  ggplot(weather.Panel, aes(interval60,Temperature, color = Temperature)) + geom_line() + 
    scale_color_gradient(low = "#74add1", high = "#f46d43")+
    labs(title="Temperature", x="Hour", y="Temperature (°F)") + theme_bw(),
  
  top="Weather Data - Philadelphia, 2022")
```

### Parks
**Question for Michael/Matt**: What's the best way to call on the Open Data Philly REST API?
```{r parks}
# parks <- st_read("https://services.arcgis.com/fLeGjb7u4uXqeF9q/arcgis/rest/services/PPR_Districts_2018/FeatureServer/0/query?outFields=*&where=1%3D1")
```

Various sources: https://openmaps.phila.gov/

### Land use
* Figure out how to access the [Open Data Philly Land Use data set](https://www.opendataphilly.org/dataset/land-use)

### Commercial corridors
https://metadata.phila.gov/#home/representationdetails/56423a4e902dbdd813db9a55/

### Bike network
```{r bike network}
#bike_network <- st_read("https://mapservices.pasda.psu.edu/server/services/pasda/PhiladelphiaBikeNetwork_SupportingDatasets/MapServer/WMSServer?request=GetCapabilities&service=WMS")

#https://metadata.phila.gov/#home/representationdetails/55438ac89b989a05172d0d77/

#https://www.pasda.psu.edu/uci/DataSummary.aspx?dataset=1026

```

### High Injury Network
```{r high injury network}
#high_injury_network <- st_read("https://phl.carto.com/api/v2/sql?q=SELECT * FROM high_injury_network_2020")

#https://www.opendataphilly.org/dataset/high-injury-network
```

### PHL City Landmarks
```{r city landmarks}

https://www.opendataphilly.org/dataset/city-landmarks
```

### Street Nodes
https://www.opendataphilly.org/dataset/street-nodes

### Street centerlines
https://www.opendataphilly.org/dataset/street-centerlines

### SEPTA locations
https://www.opendataphilly.org/dataset/septa-locations-api

# Exploratory Analysis
##Map current stations

Map stations
```{r plot stations}

ggplot()+
  geom_sf(data = blockgrps)+
  geom_sf(data = stations, color = "red") +
  # ylim(39.89, 40.01)+
  # xlim(-75.23, -75.12)+
  mapTheme
```

Map stations by ridership - rides appear to be clustered in center city
```{r stations by ridership}
ggplot()+
  geom_sf(data = blockgrps)+
  geom_sf(data=station_census, aes(size = ridership, color = ridership), alpha = 0.7)+
  ylim(39.89, 40.01)+
  xlim(-75.23, -75.12)+
  mapTheme
```

Plot distribution of trips
```{r plot trip distributions}
ggplot()+
  geom_histogram(data = station_census, aes(x = ridership))

```

Map stations by ridership quartile
```{r stations by ridership quartile}
station_census %>%
         mutate(ridership_quartile = case_when( # based on summary(station_census$ridership)
           ridership <= 1788 ~ "1st Quartile",
           ridership > 1788 & ridership <= 3919 ~ "2nd Quartile", #based on median
           ridership > 3919 & ridership <= 7105 ~ "3rd Quartile",
           ridership > 7105 ~ "4th Quartile"
         )) %>%
  ggplot(data = .)+
    geom_sf(data = blockgrps)+
    geom_sf(aes(color = ridership_quartile, size = ridership), alpha = 0.7)+
    ylim(39.89, 40.01)+
    xlim(-75.23, -75.12)+
    mapTheme
```

## Plot correlations
Plot initial correlations between census variables and station ridership - this is a very busy plot, but gives us a sense of when ridership peaks (as expected, in summer).
```{r exploratory correlations}
st_drop_geometry(station_census) %>%
  dplyr::select(ridership, pct_rent_spent, pct_white, pct_employed) %>%
  gather(Variable, Value, - ridership) %>%
  ggplot(aes(Value, ridership))+
  geom_point(size = 0.5, color = "black")+
  geom_smooth(method = lm, se = FALSE, color = "#cc4778", size = 2)+
  scale_x_continuous(n.breaks = 5)+
  facet_wrap(~Variable, ncol = 2, scales = "free")+
  labs(title = "Ridership as a function of station features")+
  plotTheme
```

Number of rides by station by week - the steep drop at week 40 correlates with bad weather in the same week.
```{r rides by station by week}
ggplot(rides)+
  geom_freqpoly(aes(week, color = start_station), binwidth = 1)
  

rides_per_station_per_week <- rides %>%
  group_by(week, start_station) %>%
  summarize(rides_per_week = length(unique(trip_id)))

ggplot(rides_per_station_per_week)+
  geom_line(aes(x=week, y=rides_per_week, color = start_station))+
  guides(color = "none")+
  labs(title = "Rides by station by week, 2022")+
  plotTheme
```

Number of rides by bike type by week
```{r rides by station by week}
ggplot(rides)+
  geom_freqpoly(aes(week, color = bike_type), binwidth = 1)+
  labs(title = "Rides by bike type over the course of 2022")+
  plotTheme
```
